
## COS-D407. *Scientific Modeling and Model Validation*    
   
#### Hands-on excercises 

#### Week 3 
  
#### University of Helsinki, Finland 

#### 26.10.2020--09.12.2020    
   
#### Lecturer: Christina Bohk-Ewald  
      
##### Source: https://github.com/christina-bohk-ewald/2020-COS-D407-scientific-modeling-and-model-validation
  
####  
  
\pagebreak  
  
## Table of content:  
   
#### 1. Some preparations in R  
  
#### 2. Download and load required input data 
  
#### 3. Estimate COVID-19 infections in Finland  
    
#### 4. Time for you to think both creatively and critically about the meaning of these results
  
####  
  
\pagebreak
  
---  
author:
  Christina Bohk-Ewald  
output: pdf_document
params:  
  date: "`r Sys.time()`"  
  printcode: TRUE  
---  
  
```{r, setup, include=FALSE}  
# set this option in the first code chunk in the document  
knitr::opts_chunk$set(echo = params$printcode)  
```  

## 1. Some preparations in R

### 1.1 Open a new script for week 3 in R (e.g., *week-3.R*) and save it to a folder of your choice (e.g., *course-COS-D407*).    
  
### 1.2 Create a filepath to this folder from where you would like to load data and to where you would like to save your outcome. For example,

```{r include=TRUE, message=FALSE,warning=FALSE, eval=FALSE}  
the.course-COS-D407.path <- c("C:/course-COS-D407")  
```  

### 1.3 You can then set the working directory to this path  
  
```{r include=TRUE, message=FALSE,warning=FALSE,eval=FALSE}  
setwd(the.course-COS-D407.path)   
```    
  
## 2. Download and load required input data   
  
In week 3 we apply the [demographic scaling model of Bohk-Ewald et al. (2020)](https://doi.org/10.1101/2020.04.23.20077719) for estimating total numbers of COVID-19 infections in Finland. We will start with downloading and loading required input data in R, continue with reading basic functions of the demographic scaling model in R, and finally applying it to estimate the number of people who probably are and have been infected with COVID-19 so far. 
  
### 2.1 Download confirmed cases and reported deaths attributable to COVID-19 as of today  
  
Please go to the website of the [Johns Hopkins University CSSE](https://data.humdata.org/dataset/novel-coronavirus-2019-ncov-cases). The files  
  
- *time_series_covid19_confirmed_global.csv* 
- *time_series_covid19_deaths_global.csv* 

contain confirmed cases and reported deaths, respectively, for many countries on a daily basis since January 22, 2020. Please download these two files and save them in your project folder.       
    
### 2.2 Load COVID-19 data  

Please load the numbers of confirmed cases and reported deaths from COVID-19 in R using the function *read.csv* of the R-package *openxlsx*.  
  
```{r include=TRUE, message=FALSE,warning=FALSE}  
require(openxlsx)

confirmed <- read.csv("time_series_covid19_confirmed_global.csv",header=TRUE, 
stringsAsFactors = FALSE)
confirmed[1:2,1:6]

deaths <- read.csv("time_series_covid19_deaths_global.csv",header=TRUE, 
stringsAsFactors = FALSE)
deaths[1:2,((ncol(deaths)-5):ncol(deaths))]

```  
  
### 2.3 Download and load abridged life tables

Please go to the [UNWPP2019 website](https://population.un.org/wpp/), [download abridged life tables for both sexes together](https://population.un.org/wpp/Download/Files/1_Indicators%20(Standard)/EXCEL_FILES/3_Mortality/WPP2019_MORT_F17_1_ABRIDGED_LIFE_TABLE_BOTH_SEXES.xlsx), save them into your project folder, and then load them into R. 

```{r include=TRUE, message=FALSE, warning=FALSE, eval=TRUE} 
  
lt_1950_2020 <- read.xlsx("WPP2019_MORT_F17_1_ABRIDGED_LIFE_TABLE_BOTH_SEXES.xlsx",
sheet = 1,startRow = 17)
  
``` 

Brief data description. The data object *lt_1950_2020* contains abridged life tables for both sexes for all UN countries.  
  
Explore this data object and find out how large Finnish remaining life expectancy at birth has been 1950-55 through 2015-19.  
  
```{r include=TRUE, message=FALSE, warning=FALSE, eval=TRUE} 
  
## lt_1950_2020[1:2,]  
  
colnames(lt_1950_2020)  
  
lt_1950_2020[which(lt_1950_2020[,"Region,.subregion,.country.or.area.*"]=="Finland" & 
  lt_1950_2020["Age.(x)"]==0),c("Period","Expectation.of.life.e(x)")]
  
```   
    
### 2.4 Load global pattern over age of COVID-19 deaths

[Dudel et al. (2020)](https://doi.org/10.1371/journal.pone.0238904) provide in their supplementary material data for age-specific death counts attributable to COVID-19, which has been served as a basis for calculating a global average pattern over age for total death counts as input for the demographic scaling model. You can download this *global average pattern over age* from the [GitHub repository for this course](https://github.com/christina-bohk-ewald/2020-COS-D407-scientific-modeling-and-model-validation).     
  
```{r include=TRUE, message=FALSE, warning=FALSE, eval=TRUE} 

global_age_dist_deaths <- source("global_age_dist_deaths.R")
## global_age_dist_deaths
  
``` 

Brief data description. The data object *global_age_dist_deaths* contains the global pattern over 10-year age groups of COVID-19 deaths.
  
### 2.5 Load infection fatality rates from Verity et al. (2020)

[Verity and colleagues (2020, page 5)](https://doi.org/10.1016/S1473-3099(20)30243-7) report infection fatality rates by 10-year age groups for Hubei, China, on page 5. Please create a data object *ifr_by_age_china_verity* that contains these data or download it from the [GitHub repository for this course](https://github.com/christina-bohk-ewald/2020-COS-D407-scientific-modeling-and-model-validation).
  
```{r include=TRUE, message=FALSE, warning=FALSE, eval=TRUE} 

ifr_by_age_china_verity <- read.table("infection-fatality-rates-by-age-china-Verity.txt",
header=FALSE, stringsAsFactors = FALSE)

ifr_by_age_china_verity
  
``` 
  
Brief data description. The data object *ifr_by_age_china_verity* contains the modal estimate as well as the lower and upper bound of the 95 percent credible interval for the infection fatality rates of Hubei, China, by 10-year age groups.   
  
## 3. Estimate COVID-19 infections in Finland  
  
### 3.1 Source basic functions of the demographic scaling model

You can find the basic functions of the demographic scaling model in the file *basic-functions-week-3.R* in the [GitHub repository for this course](https://github.com/christina-bohk-ewald/2020-COS-D407-scientific-modeling-and-model-validation). They contain the functions:   

- *to_ungroup* to interpolate IFR estimates of Verity et al. (2020) into single years of age
- *get_ungrouped_ex_2015_2020* to ungroup remaining life expectancy
- *map_ifr_betw_ref_and_one_coi_thanatAge* to scale IFRs from a reference country (here: China; Verity et al. (2020)) onto a country of interest based on remaining life expectancy
- *aggregate_mapped_ifr_10y* to aggregate scaled IFRs into 10-year age groups
- *disaggregate_deaths_one_coi_10y* to disaggregate total deaths into 10-year age groups    

You may have a look at these basic functions if you wish. But it is also fine to just source (or load) them via the file *basic-functions-week-3.R*:

```{r include=TRUE, message=FALSE, warning=FALSE, eval=TRUE} 

source("basic-functions-week-3.R")  

```  
  
### 3.2 Apply the demographic scaling model with data for Finland  
  
```{r include=TRUE, message=FALSE, warning=FALSE, eval=TRUE} 

#
## 1. Ungroup China's IFR:
#

ungrouped_mode_ifr_by_single_age_china_sp <- to_ungroup(to_ungroup=
				ifr_by_age_china_verity[,2],nr_grouped_years=10)

ungrouped_low95_ifr_by_single_age_china_sp <- to_ungroup(to_ungroup=
				ifr_by_age_china_verity[,3],nr_grouped_years=10)

ungrouped_up95_ifr_by_single_age_china_sp <- to_ungroup(to_ungroup=
				ifr_by_age_china_verity[,4],nr_grouped_years=10)

#
## 2. Scale IFRs from a RC onto a COI via remaining life expectancy:
#

mapped_mode_ifr_thanatAge <- map_ifr_betw_ref_and_one_coi_thanatAge(coi="Finland",
	lt_1950_2020=lt_1950_2020,
	ungrouped_ifr_by_single_age_china_sp=ungrouped_mode_ifr_by_single_age_china_sp)

## and fill in the few NA values:

pos_na <- which(is.na(mapped_mode_ifr_thanatAge[1,]))
	if(length(pos_na)>0){
		for(pos in 1:length(pos_na)){
			if(pos_na[pos] < 6){
				mapped_mode_ifr_thanatAge[1,pos_na[pos]] <- 
					min(mapped_mode_ifr_thanatAge[1,],na.rm=TRUE)
			}
			if(pos_na[pos] >= 6){
				mapped_mode_ifr_thanatAge[1,pos_na[pos]] <- 
					mapped_mode_ifr_thanatAge[1,pos_na[pos]-1]
			}
		} ## for pos
	} ## if

#
## 3. Put scaled IFRs into 10-year age groups: 
#

mapped_mode_ifr_thanatAge_10y <- aggregate_mapped_ifr_10y(disaggregated_mapped_ifr=
						mapped_mode_ifr_thanatAge)

#
## 4. Disaggregate total COVID-19-related deaths into 10-year age groups:
#

deaths_by_age <- disaggregate_deaths_one_coi_10y(coi="Finland")
	
#
## 5. Estimate COVID-19 infections over time:
#

inf_mode <- colSums( deaths_by_age / mapped_mode_ifr_thanatAge_10y ) 

```  
  
Think about how you could extend this R-code in order to estimate the total numbers of COVID-19 infections for other countries and how to provide uncertainty estimates for them.  
  
### 3.3 Visualize COVID-19 infecton estimates for Finland  
  
Visualize the numbers of COVID-19 infections in Finland over time, accounting for an average time to death of 18 days.  
   
```{r include=TRUE, message=FALSE, warning=FALSE, eval=TRUE} 
  
dates <- seq(as.Date("22/01/2020", format = "%d/%m/%Y"),
by = "days", length = (ncol(deaths)-4) )  
  
```  
  
```{r fig.pos="h", fig.height=7, fig.width=10, fig.align="center", include=TRUE, message=FALSE, warning=FALSE}   
  
par(fig = c(0,1,0,1), las=1, mai=c(0.4,0.8,0.8,0.4))

plot(x=-100,y=-100,xlim=c(0,length(5:ncol(deaths))),ylim=c(0,20),xlab="Date",ylab="",cex.main=0.9,
main="Total numbers of COVID-19 infections, in thousand, in Finland",axes=FALSE)

segments(x0=rep(0,4),x1=rep(length(5:ncol(deaths)),4),y0=seq(5,20,5),y1=seq(5,20,5),
lty=2,col=grey(0.8))

lines(x=1:length(5:ncol(deaths)),y=c(inf_mode[-c(1:18)],rep(NA,18))/1000,col="blue",lwd=3)

lines(x=1:length(5:ncol(deaths)),y=confirmed[which(confirmed[,"Country.Region"]=="Finland"),
5:ncol(confirmed)]/1000,col="black",lty=2,lwd=3)

legend(0,20,c("Estimated","Confirmed"),col=c("blue","black"),bty="n",lwd=2,lty=1)

axis(side=1,at=seq(1,length(5:ncol(deaths)),7),labels=FALSE,lwd=1,pos=0)

axis(side=1,at=c(seq(1,length(5:ncol(deaths)),21),length(5:ncol(deaths))),
labels=dates[c(seq(1,length(5:ncol(deaths)),21),
length(5:ncol(deaths)))],lwd=3,pos=0)

axis(side=2,at=seq(0,20,1),labels=FALSE,lwd=1,pos=0)

axis(side=2,at=seq(0,20,5),labels=TRUE,lwd=3,pos=0)
  
```  

Please describe the level and the temporal development of the estimated numbers of COVID-19 infections in Finland, also compared to (1) the numbers of confirmed cases in Finland, and to (2) the corresponding figures in other European countries. As a source of inspiration, you may want to have a look at the [press release for the paper of Bohk-Ewald et al. (2020) on *A demographic scaling model for estimating the total number of COVID-19 infections*](https://www.helsinki.fi/en/news/society-economy/how-many-finns-have-really-been-infected-with-covid-19). 
  
Again, please think about how you could extend this R-code in order to estimate the total numbers of COVID-19 infections for other countries and how to provide uncertainty estimates for them.  
  
## 4. Time for you to think both creatively and critically about these COVID-19 infection estimates for Finland. How plausible are they, also considering by how much the key assumptions of the demographic scaling model may hold?
     
As a source of inspiration, you may want to have a look at the paper of Bohk-Ewald et al. (2020) on [*A demographic scaling model for estimating the total number of COVID-19 infections* with a preprint being available on medRxiv](https://doi.org/10.1101/2020.04.23.20077719). 


 